---
title: "Section9_datacleaning/wrangling"
format: html
editor: visual
---

# Data Resources 

## Read in the data

```{r}
library(dplyr)
library(tidyr)
library(readr)
```

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

```

```{r}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"))

```

## Explore the Data

```{r}
## Prints the column names of my data frame
colnames(catch_original)

## First 6 lines of the data frame
head(catch_original)

## Summary of each column of data
summary(catch_original)

## Prints unique values in a column (in this case, the region)
unique(catch_original$Region)

## Opens data frame in its own tab to see each row and column of the data (do in console)
View(catch_original)
```

# Clean and Reshape data

## Using `select()` function

```{r}
catch_data <- catch_original %>%
    select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)
```

```{r}
catch_data <- catch_original %>%
    select(-All,-notesRegCode)
```

## Check column types

```{r}
dplyr::glimpse(catch_data)
```

## Replace values in a column with `mutate()`

```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = as.numeric(Chinook))
```

```{r}
head(catch_clean)
i <- which(is.na(catch_clean$Chinook))
i
```

```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = if_else(condition = Chinook == "I", 
                             true = "1", 
                             false = Chinook),
           Chinook = as.numeric(Chinook))

##check
catch_clean[i, ]
```

## Reshape data with `pivot_longer()` and `pivot_wider()`

```{r}
catch_long <- catch_clean %>% 
    #pivot longer all columns except Region and Year
    pivot_longer(
        cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch"
    )

head(catch_long)
```

```{r}
catch_wide <- catch_long %>%
    pivot_wider(names_from = species,
                values_from = catch)

head(catch_wide)
```

## Rename columns `rename()`

```{r}
catch_long <- catch_long %>%
    rename(catch_thousands = catch)

head(catch_long)
```

## Add columns with `mutate()`

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000)

head(catch_long)
```

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000) %>%
    select(-catch_thousands)

head(catch_long)
```

## Summary stats using `group_by()` and `summarize()`

```{r}
unique(catch_original$Region)
```

```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch))

head(mean_region)
```

```{r}
n_region <- catch_long %>%
    group_by(Region) %>%
    summarize(n = n())

head(n_region)
```

```{r}
## for example:
catch_year_sp <- catch_long %>%
    group_by(Year, species) %>%
    summarize(total_year = sum(catch, na.rm = T))
```

## Filtering rows using `filter()`

```{r}
sse_catch <- catch_long %>%
    filter(Region == "SSE")

head(sse_catch)
```

```{r}
## Catches over a million fish
catch_million <- catch_long %>%
    filter(catch > 1000000)

## Chinook from SSE data
chinook_see <- catch_long %>%
    filter(Region == "SSE",
           species == "Chinook")

## OR
chinook_see <- catch_long %>%
    filter(Region == "SSE" & species == "Chinook")
```

## Sort data using `arrange()`

```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(mean_catch)

head(mean_region)
```

```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(desc(mean_catch))

head(mean_region)
```

## Split and combine values in columns with `separate()` and `unite()`

```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAU-039"))

sites_df %>%
    separate(site, c("island", "site_number"), "-")
```

```{r}
## create `cities_df`
cities_df <- data.frame(city = c("Juneau AK",
                                 "Sitka AK",
                                 "Anchorage AK"))
```

```{r}
dates_df <- data.frame(
    year = c("1930",
             "1930",
             "1930"),
    month = c("12",
              "12",
              "12"),
    day = c("14",
            "15",
            "16")
)

dates_df %>%
    unite(date, year, month, day, sep = "-")
```

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```
